A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-22 tidyindex.tex
small changes suggested by Nicolas

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Examples of using a pipeline approach for data analysis can be traced
back to the interactive graphics literature, including A. Buja et al.
(1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
Wickham et al. (2009). Wickham et al. (2009) argue that whether made
explicit or not, a pipeline has to be presented in every graphics
program, and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of interactive graphics programs, it
is also applicable generally to any data analysis workflow. More
recently, the tidyverse suite (Wickham et al. 2019) takes the pipeline
approach for general-purpose data wrangling and has gained popularity
within the R community. The pipeline-style code can be directly read as
a series of operations applied successively on tidy data objects,
offering a method to document the data wrangling process with all the
computational details for reproducibility.

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

Variable transformation takes the input of a single variable
\(x_j(s;t)\) and reshapes its distribution using the function \(T[.]\)
to produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
distribution, transformations such as log, square root, or cubic root
can adjust the distribution towards normality. For example, in Human
Development Index (HDI), a logarithmic transformation is applied to the
variable Gross National Income per capita (GNI), to reduce its impact on
HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardization, and min-max standardization and can be expressed as
\([x_{j}(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
three dimensions (health, education, and economy) are converted into the
same scale (0-1) using min-max standardization.

@@

Simplification takes a continuous variable \(x_j(s;t)\) and categorises
it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
through a piecewise function,

@@

The output contains the original data, index values (\texttt{.index}),
parameters used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
\texttt{.fitted}). This data can be visualized to investigate the
spatio-temporal distribution of the drought/flood events, as well as the
response of index values to different time scales and distribution
parameters at specific single locations.
Figure~\ref{fig-compute-spatial} and Figure~\ref{fig-compute-temporal}
exemplify two possibilities. Figure~\ref{fig-compute-spatial} presents
the spatial distribution of SPI during two periods: October 2010 to
March 2011 for the 2010/11 Queensland flood and October 2019 to March
2020 for the 2019 Australia drought, which contributes to the notorious
2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
sensitivity of the SPEI series at the Texas post office to different
time scales and fitted distributions. Larger time scales produce a
smoother index across time, however, all time scales indicate an extreme
drought (corresponding to -2 in SPEI) in 2020, confirming the severity
of the drought across different time horizons. Moreover, the chosen
distribution has less influence on the index, with general extreme value
distribution tending to produce more extreme outcomes than log-logistic
distribution for the extreme events (index \textgreater{} 2 or
\textless-2).

@@

In SPI, the distribution fit step fits the gamma distribution to the
aggregated precipitation series separately for each month. This results
in 31 or 32 points, from January 1990 to April 2022, for estimating each
set of distribution parameters. To account for this sampling uncertainty
with these samples, bootstrapping is used to generate replicates of the
aggregated series. In the \texttt{tidyindex} package, this bootstrap
sampling is activated when the argument \texttt{.n\_boot} is set to a
value other than the default of 1. In the following code, the
Standardized Precipitation Index (SPI) is calculated using a time scale
of 24. The bootstrap procedure samples the aggregated precipitation
(\texttt{.agg}) for 100 iterations (\texttt{.n\_boot\ =\ 100}) and then
fits the gamma distribution. The resulting gamma probabilities are then
transformed into normal densities in the normalizing step with
\texttt{agument()}.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. The relatively wide confidence
interval, as well as during the excessive precipitation events in
1996-1998 and 1999-2000, suggests a high variation of the gamma
parameters estimated from the bootstrap samples and its difficulty to
accurately quantifying the drought and flood severity in extreme events.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

Examples of using a pipeline approach for data analysis can be traced
back to the interactive graphics literature, including A. Buja et al.
(1988); Sutherland et al. (2000); Wickham et al. (2009); Xie, Hofmann,
and Cheng (2014). Wickham et al. (2009) argue that whether made explicit
or not, a pipeline has to be presented in every graphics program, and
making them explicit is beneficial for understanding the implementation
and comparing between different graphic systems. While this comment is
made in the context of interactive graphics programs, it is also
applicable generally to any data analysis workflow. More recently, the
tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
general-purpose data wrangling and has gained popularity within the R
community. The pipeline-style code can be directly read as a series of
operations applied successively on tidy data objects, offering a method
to document the data wrangling process with all the computational
details for reproducibility.

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
\bottomrule()
\end{longtable}

@@

Variable transformation takes the input of a single variable
\(x_j(s;t)\) and reshapes its distribution using the function \(T[.]\)
to produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
distribution, transformations such as log, square root, or cubic root
can adjust the distribution towards normality. For example, in the Human
Development Index (HDI), a logarithmic transformation is applied to the
variable Gross National Income per capita (GNI), to reduce its impact on
HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardization, and min-max standardization and can be expressed as
\([x_{j}(s;t) - \alpha]/\gamma\). In the Human Development Index (HDI),
the three dimensions (health, education, and economy) are converted into
the same scale (0-1) using min-max standardization.

@@

Simplification takes a continuous variable \(x_j(s;t)\) and categorises
it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\)
through a piecewise constant function,

@@

The output contains the original data, index values (\texttt{.index}),
parameters used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
\texttt{.fitted}). This data can be visualized to investigate the
spatio-temporal distribution of the drought/flood events, as well as the
response of index values to different time scales and distribution
parameters at specific single locations.
Figure~\ref{fig-compute-spatial} and Figure~\ref{fig-compute-temporal}
exemplify two possibilities. Figure~\ref{fig-compute-spatial} presents
the spatial distribution of SPI during two periods: October 2010 to
March 2011 for the 2010/11 Queensland flood and October 2019 to March
2020 for the 2019 Australia drought, which contributes to the notorious
2019/20 bushfire season. Figure~\ref{fig-compute-temporal} displays the
sensitivity of the SPEI series at the Texas post office to different
time scales and fitted distributions. Larger time scales produce a
smoother index across time, however, all time scales indicate an extreme
drought (corresponding to -2 in SPEI) in 2020, confirming the severity
of the drought across different time horizons. Moreover, the chosen
distribution has less influence on the index, with general extreme value
distribution tending to produce more extreme outcomes than log-logistic
distribution for the extreme events (index \textgreater{} 2 or
\textless-2).

@@

In SPI, the distribution fit step fits the gamma distribution to the
aggregated precipitation series separately for each month. This results
in 31 or 32 points, from January 1990 to April 2022, for estimating each
set of distribution parameters. To account for this sampling uncertainty
with these samples, bootstrapping is used to generate replicates of the
aggregated series. In the \texttt{tidyindex} package, this bootstrap
sampling is activated when the argument \texttt{.n\_boot} is set to a
value other than the default of 1. In the following code, the
Standardized Precipitation Index (SPI) is calculated using a time scale
of 24. The bootstrap procedure samples the aggregated precipitation
(\texttt{.agg}) for 100 iterations (\texttt{.n\_boot\ =\ 100}) and then
fits the gamma distribution. The resulting gamma probabilities are then
transformed into normal densities in the normalizing step with
\texttt{augment()}.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. The relatively wide confidence
interval, as well as during the excessive precipitation events in
1996-1998 and 1999-2000, suggests a high variation of the gamma
parameters estimated from the bootstrap samples and its difficulty to
accurately quantify the drought and flood severity in extreme events.
